---
title: "Solution for the exercise on PCA"
author: "Radchuk Viktoriia"
date: "2022-03-20"
output:
  html_document:
    theme: default
       
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(vegan)
```

# Read in the data
```{r LH data read, echo = TRUE}
#  reading in the file with the 143 species of interest
spSEAsia <- read.csv(file = here::here('data', 'Species_SEAsia_02_02_16_fin.csv'))
str(spSEAsia)

# read in the PanTheria database
pant <- read.delim(file = here::here('data', 'Pantheria.txt'))
str(pant)
```

# Prepare data    
```{r data prep, echo = TRUE}
## joining both datasets together
pant$Sp_bin <- pant$MSW05_Binomial
LH <- merge(pant, spSEAsia, by = "Sp_bin")
nrow(LH)  

## Select 5 life history traits of interest
LH_subs <- LH %>%
  select('Genus', 'Species', 'Sp_bin', 
         'X5.1_AdultBodyMass_g',  'X15.1_LitterSize',
         'X6.1_DietBreadth', 'X22.1_HomeRange_km2',
         'X27.2_HuPopDen_Mean_n.km2', 'X28.1_Precip_Mean_mm',
         'X28.2_Temp_Mean_01degC', 'X30.1_AET_Mean_mm')


# replacing -999 with NA
LH_subs[LH_subs == -999] <-  NA

LH_noNA <- na.omit(LH_subs)
names(LH_noNA) <- sub("^[^_]*_", "", names(LH_noNA))

str(LH_noNA)
```

# Run PCA on environmental variables
```{r run PCA, echo=TRUE}
env_LH <- LH_noNA %>% 
  select('HuPopDen_Mean_n.km2', 'Precip_Mean_mm',
         'Temp_Mean_01degC', 'AET_Mean_mm')

pca_env_SEAs <- rda(env_LH, scale = TRUE)
```
Since our variables (human population density, precipitation, temperature...) are in very different units, we use option 'scale=', so that all variables are now unitless and comparable.

# Interpreting and visualizing results
```{r interpret PCA, echo = TRUE, fig.align='center',out.height='60%', out.width='60%'}
summary(pca_env_SEAs)

# extract eigenvalues
eigval <- pca_env_SEAs$CA$eig
eigval[eigval > mean(eigval)]  

bstick(pca_env_SEAs); eigval

biplot(pca_env_SEAs, scaling = 'species')
```
So the first principal component should be sufficient but since we can easily plot in 2D, we will also extract the second one (and focus in interpretting on the pattern along the PC1).


```{r nicer-plot, out.height='80%', out.width='80%', fig.align='center'}
var_PCA <- scores(pca_env_SEAs, choices = c(1,2), scaling = 2, tidy = TRUE)

LH_noNA$label <- rownames(LH_noNA)
vis_PCA <- merge(subset(var_PCA, score =='sites'), subset(LH_noNA, select = c('label', 'bin')), by = 'label')

spec_plot <- var_PCA %>% 
  filter(score == 'species')
ggplot(vis_PCA, aes(x = PC1, y = PC2)) +
   geom_text(aes(label = bin), check_overlap = TRUE) +
  geom_segment(data = spec_plot, 
               aes(x=0, y=0, xend=PC1, yend=PC2), 
               arrow = arrow(length=unit(.5, 'cm')),
               col ='red') +
  geom_text(data = spec_plot, aes(label = label),             nudge_y = -0.1, col = 'red') +
theme_bw() +
  geom_hline(yintercept = 0, lty = 3) +
  geom_vline(xintercept = 0, lty = 3)
```




